---
title: "Environment Setup with uv and reticulate"
keep-md: true
execute:
  cache: true
  freeze: auto
  warning: false
  message: false
---

## Philosophy: Reproducible Python Environments

GeneSelectR uses **uv** for Python dependency management instead of traditional `pip` + `virtualenv` because:

1. **Speed**: uv is written in Rust and is 10-100× faster than pip
2. **Determinism**: Lockfile-style resolution ensures reproducible installs
3. **Simplicity**: Single tool for environment creation and package management

The R-Python bridge is handled by **reticulate**, which allows seamless calling of Python objects and functions from R.

## Setting Up the Python Environment

### Step 1: Install uv (if not already installed)

```bash
# macOS/Linux
curl -LsSf https://astral.sh/uv/install.sh | sh

# Or via Homebrew
brew install uv
```

### Step 2: Create a Virtual Environment

From the GeneSelectR project root:

```bash
cd /Users/zero/Desktop/GeneSelectR
uv venv .venv
```

This creates a `.venv/` directory containing an isolated Python installation.

### Step 3: Install Python Dependencies

GeneSelectR's Python requirements are specified in `inst/python/requirements.txt`:

```bash
uv pip install -r inst/python/requirements.txt
```

This installs:

- `scikit-learn`: Core ML algorithms
- `numpy`, `pandas`: Data manipulation
- `scikit-optimize`: Bayesian hyperparameter optimization
- `boruta`: Boruta feature selection
- Additional dependencies as needed

### Step 4: Bind reticulate to the uv Environment

In R, point reticulate to the uv-managed Python interpreter:

```{r setup-reticulate}
library(tidyverse)
library(reticulate)

# Point to the uv venv
project_root <- "/Users/zero/Desktop/GeneSelectR"
venv_path <- file.path(project_root, ".venv")
Sys.unsetenv("RETICULATE_PYTHON")
use_python(file.path(venv_path, "bin", "python"), required = TRUE)

# Verify configuration
py_config()
```

**Output** (example):

```
python:         /Users/zero/Desktop/GeneSelectR/.venv/bin/python
libpython:      /Users/zero/Desktop/GeneSelectR/.venv/lib/libpython3.11.dylib
pythonhome:     /Users/zero/Desktop/GeneSelectR/.venv
version:        3.11.5
numpy:          /Users/zero/Desktop/GeneSelectR/.venv/lib/python3.11/site-packages/numpy
sklearn:        /Users/zero/Desktop/GeneSelectR/.venv/lib/python3.11/site-packages/sklearn
```

### Step 5: Test the Connection

```{r test-import, eval=FALSE}
# Import Python modules via reticulate
sklearn <- import("sklearn")
np <- import("numpy")

# Check versions
cat("scikit-learn version:", sklearn$`__version__`, "\n")
cat("numpy version:", np$`__version__`, "\n")
```

If imports succeed without errors, the environment is ready.

## Reproducibility Guarantees

### Cache and Freeze

Quarto's `freeze: auto` setting ensures:

- Computation results are cached after first render
- Re-renders only happen when source `.qmd` changes
- Prevents drift from upstream package updates during documentation updates

### Random Seeds

For reproducible machine learning:

```{r set-seeds, eval=FALSE}
set.seed(42)  # R seed
py_run_string("import numpy as np; np.random.seed(42)")  # Python seed
```

Pass `random_state` parameters to sklearn functions explicitly.

### Environment Snapshots

To capture exact package versions:

```bash
uv pip freeze > requirements.lock
```

Restore later with:

```bash
uv pip install -r requirements.lock
```

## Design Rationale

### Why Not Conda/Mamba?

While Conda is popular in bioinformatics, uv offers:

- **Lighter weight**: No 500MB+ base installation
- **Faster installs**: Parallel downloads and installs
- **Better R integration**: Standard Python venvs work seamlessly with reticulate

### Why Not renv for Python?

R's `renv` can manage Python, but:

- Less mature than uv for Python-specific workflows
- Slower package resolution
- Limited support for scikit-learn's binary dependencies

### Why Separate from Package Installation?

GeneSelectR does **not** bundle Python dependencies in the R package because:

1. **CRAN policy**: Binary Python packages would bloat package size
2. **Flexibility**: Users may have existing ML environments
3. **Isolation**: Documentation environment is independent of package tests

## Troubleshooting

### "Python module not found"

```r
# Check module availability
py_module_available("sklearn")

# List installed packages
py_run_string("import pkg_resources; print([p.project_name for p in pkg_resources.working_set])")
```

### "Wrong Python version"

Ensure `use_python()` points to the uv venv:

```r
Sys.which("python")  # Should NOT be system Python
py_config()$python   # Should be /path/to/.venv/bin/python
```

### "Segmentation fault on import"

Likely a binary compatibility issue. Rebuild the environment:

```bash
rm -rf .venv
uv venv .venv
uv pip install -r inst/python/requirements.txt
```

## Summary

With uv and reticulate configured:

1. Python dependencies are isolated in `.venv/`
2. R can call Python functions via `import()` and `py_run_*()`
3. Reproducibility is ensured via lockfiles and caching
4. No modifications to GeneSelectR source code are needed

---

**Next**: [Core Classes: S4 Objects and Semantics](03_core_classes.qmd)
